---
title: "Lab 6: Machine Learning in Hydrology"
subtitle: 'Ecosystem Science and Sustainability 330'
author:
- name: Natasha Zizic
email: "nzizic@colostate.edu"
format: html
outtput: docs 
execute:
  echo: true
---

### Lab Set Up

```{r}
library(tidyverse)
library(tidymodels)
library(powerjoin)
library(glue)
library(vip)
library(baguette)
library(recipes)
```

### Question 1:

```{r}
root  <- 'https://gdex.ucar.edu/dataset/camels/file'
```

```{r}
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf', 
              'data/camels_attributes_v2.0.pdf')
```

```{r}
types <- c("clim", "geol", "soil", "topo", "vege", "hydro")
```

```{r}
remote_files  <- glue('{root}/camels_{types}.txt')

local_files   <- glue('data/camels_{types}.txt')
```

```{r}
walk2(remote_files, local_files, download.file, quiet = TRUE)
```

```{r}
camels <- map(local_files, read_delim, show_col_types = FALSE) 
```

```{r}
camels <- power_full_join(camels ,by = 'gauge_id')
```

In the camels dataset, zero_q_frequency represents the days in which there is no water flow. This is when Q = 0 mm / day.

### Question 2:

```{r}
ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = aridity)) +
  scale_color_gradient(low = "dodgerblue", high = "orange") +
  ggthemes::theme_map()
```

```{r}
ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = p_mean)) +
  scale_color_gradient(low = "orange", high = "dodgerblue") +
  ggthemes::theme_map()
```

### Question 3:

```{r}
camels |> 
  select(aridity, p_mean, q_mean) |> 
  drop_na() |> 
  cor()
```

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm", color = "red", linetype = 2) +
  scale_color_viridis_c() +
  theme_linedraw() + 
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runnoff", 
       x = "Aridity", 
       y = "Rainfall",
       color = "Mean Flow")
```

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  scale_color_viridis_c() +
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runnoff", 
       x = "Aridity", 
       y = "Rainfall",
       color = "Mean Flow")
```

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  scale_color_viridis_c(trans = "log") +
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2.5, "cm"),
        legend.key.height = unit(.5, "cm")) + 
  labs(title = "Aridity vs Rainfall vs Runnoff", 
       x = "Aridity", 
       y = "Rainfall",
       color = "Mean Flow") 
```

```{r}
set.seed(123)

camels <- camels |> 
  mutate(logQmean = log(q_mean))

camels_split <- initial_split(camels, prop = 0.8)
camels_train <- training(camels_split)
camels_test  <- testing(camels_split)

camels_cv <- vfold_cv(camels_train, v = 10)
```

```{r}

```

```{r}

```

```{r}
rec <-  recipe(logQmean ~ aridity + p_mean, data = camels_train) %>%

  step_log(all_predictors()) %>%

  step_interact(terms = ~ aridity:p_mean) |> 

  step_naomit(all_predictors(), all_outcomes())
```

```{r}

baked_data <- prep(rec, camels_train) |> 
  bake(new_data = NULL)


lm_base <- lm(logQmean ~ aridity * p_mean, data = baked_data)
summary(lm_base)
```

```{r}
summary(lm(logQmean ~ aridity + p_mean + aridity_x_p_mean, data = baked_data))
```

```{r}
test_data <-  bake(prep(rec), new_data = camels_test)
test_data$lm_pred <- predict(lm_base, newdata = test_data)
test_data <-  bake(prep(rec), new_data = camels_test)
test_data$lm_pred <- predict(lm_base, newdata = test_data)
```

```{r}
metrics(test_data, truth = logQmean, estimate = lm_pred)
```

```{r}
ggplot(test_data, aes(x = logQmean, y = lm_pred, colour = aridity)) +
  scale_color_gradient2(low = "brown", mid = "orange", high = "darkgreen") +
  geom_point() +
  geom_abline(linetype = 2) +
  theme_linedraw() + 
  labs(title = "Linear Model: Observed vs Predicted",
       x = "Observed Log Mean Flow",
       y = "Predicted Log Mean Flow",
       color = "Aridity")
```

```{r}
lm_model <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

lm_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(lm_model) %>%
  fit(data = camels_train) 

summary(extract_fit_engine(lm_wf))$coefficients
```

```{r}
lm_data <- augment(lm_wf, new_data = camels_test)
dim(lm_data)
```

```{r}
metrics(lm_data, truth = logQmean, estimate = .pred)
```

```{r}
ggplot(lm_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```

```{r}
library(baguette)
rf_model <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_model) %>%
  fit(data = camels_train) 
```

```{r}
rf_data <- augment(rf_wf, new_data = camels_test)
dim(rf_data)
```

```{r}
metrics(rf_data, truth = logQmean, estimate = .pred)
```

```{r}
ggplot(rf_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```

```{r}
wf <- workflow_set(list(rec), list(lm_model, rf_model)) %>%
  workflow_map('fit_resamples', resamples = camels_cv) 

autoplot(wf)
```

```{r}
rank_results(wf, rank_metric = "rsq", select_best = TRUE)
```

```{r}
xgb_model <- boost_tree() %>%
  set_engine("xgboost") %>%
  set_mode("regression")

```

```{r}
nn_model <- bag_mlp() %>%
  set_engine("nnet") %>%
  set_mode("regression")

```

```{r}
wf_new <- workflow_set(
  list(rec), 
  list(lm_model, rf_model, xgb_model, nn_model)
) %>%
  workflow_map('fit_resamples', resamples = camels_cv)
```

```{r}
autoplot(wf_new)
```

```{r}
rank_results(wf_new, rank_metric = "rsq", select_best = TRUE)

```

Based on this ranking, we can see that the neural network model ranked above the others. Because of this, I would choose this model to move forward with.

### Question 4:

```{r}
camels |> 
  select(high_prec_freq, low_prec_freq, q_mean) |> 
  drop_na() |> 
  cor()
```

**Data Splitting:**

```{r}
ggplot(camels, aes(x = high_prec_freq, y = low_prec_freq)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm", color = "red", linetype = 2) +
  scale_color_viridis_c() +
  theme_linedraw() + 
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runnoff", 
       x = "high precipitation frequency", 
       y = "low precipitation frequency",
       color = "Mean Flow")
```

```{r}
set.seed(275)

camels_split2 <- initial_split(camels, prop = 0.8)
camels_train2 <- training(camels_split2)
camels_test2  <- testing(camels_split2)

camels_cv2 <- vfold_cv(camels_train2, v = 10)
```

**Recipe:**

```{r}
rec2 <-  recipe(q_mean ~ high_prec_freq + low_prec_freq, data = camels_train2) %>%
  step_interact(terms = ~ high_prec_freq:low_prec_freq) |> 
  step_naomit(all_predictors(), all_outcomes())
```

```{r}
baked_data <- prep(rec2, camels_train2) |> 
  bake(new_data = NULL)

lm_base <- lm(q_mean ~ high_prec_freq * low_prec_freq, data = baked_data)
summary(lm_base)
```

```{r}
test_data2 <-  bake(prep(rec2), new_data = camels_test2)
test_data2$lm_pred2 <- predict(lm_base, newdata = test_data2)
```

```{r}
metrics(test_data2, truth = q_mean, estimate = lm_pred2)
```

```{r}
ggplot(test_data2, aes(x = q_mean, y = lm_pred2, colour = high_prec_freq)) +
  scale_color_gradient2(low = "brown", mid = "orange", high = "darkgreen") +
  geom_point() +
  geom_abline(linetype = 2) +
  theme_linedraw() + 
  labs(title = "Linear Model: Observed vs Predicted",
       x = "Observed Mean Flow",
       y = "Predicted Mean Flow",
       color = "High Precipitation Frequency")
```

**Workflow:**

```{r}
lm_model2 <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

lm_wf2 <- workflow() %>%
  add_recipe(rec2) %>%
  add_model(lm_model2) %>%
  fit(data = camels_train2) 

summary(extract_fit_engine(lm_wf2))$coefficients
```

```{r}
summary(lm_base)$coefficients
```

```{r}
lm_data2 <- augment(lm_wf2, new_data = camels_test2)
dim(lm_data)
```

```{r}
metrics(lm_data2, truth = q_mean, estimate = .pred)
```

```{r}
ggplot(lm_data2, aes(x = q_mean, y = .pred, colour = high_prec_freq)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```

```{r}
library(baguette)
rf_model2 <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_wf2 <- workflow() %>%
  add_recipe(rec2) %>%
  add_model(rf_model2) %>%
  fit(data = camels_train2) 
```

```{r}
rf_data2 <- augment(rf_wf2, new_data = camels_test2)
dim(rf_data2)
```

```{r}
metrics(rf_data2, truth = q_mean, estimate = .pred)
```

```{r}
ggplot(rf_data2, aes(x = q_mean, y = .pred, colour = high_prec_freq)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```

```{r}
wf2 <- workflow_set(list(rec2), list(lm_model2, rf_model2)) %>%
  workflow_map('fit_resamples', resamples = camels_cv2) 

autoplot(wf2)
```

```{r}
rank_results(wf2, rank_metric = "rsq", select_best = TRUE)
```

```{r}
xgb_model2 <- boost_tree() %>%
  set_engine("xgboost") %>%
  set_mode("regression")
```

```{r}
nn_model <- bag_mlp() %>%
  set_engine("nnet") %>%
  set_mode("regression")
```

```{r}
wf_new2 <- workflow_set(
  list(rec), 
  list(lm_model, rf_model, xgb_model, nn_model)
) %>%
  workflow_map('fit_resamples', resamples = camels_cv2)
```

```{r}
autoplot(wf_new2)
```

**Evaluation:**

```{r}
rank_results(wf_new2, rank_metric = "rsq", select_best = TRUE)
```

The random forest model seems to be the best one. This is because it has the highest r squared value, meaning the correlation is the strongest between variables.
